---
title: "Group 8 Combined"
author: "Ethne, Jeeya, Jolisia, Rachel, Sophia"
date: "2026-05-05"
output:
  word_document: default
  pdf_document: default
---

I. Descriptive Statistics - Rachel
```{r}
nba <- read.csv("/Users/jeeya/Downloads/TeamSummaries_raw.csv")
head(nba)
mean(nba$f_tr)
mean(nba$f_tr)
mean(nba$x3p_ar)
#plot(nba$f_tr ~ nba$f_tr, nba$x3p_ar, data = nba)
#hist(nba$f_tr ~ nba$f_tr, nba$x3p_ar, data = nba)
plot(nba$f_tr)
barplot(nba$f_tr)
plot(nba$x3p_ar)
hist(nba$x3p_ar)
hist(nba$f_tr)
mean(nba$n_rtg)
#print(mean(na.omit(nba$n_rtg))
mode(nba$n_rtg)
mean(nba$n_rtg)
nrtga <- na.omit(nba$n_rtg)
median(nrtga)
mean(nrtga)
hist(nba$n_rtg, main="Distribution of Net Rating", xlab="Net Rating", col="pink")
plot(nba$f_tr, nba$n_rtg, main="",
xlab="Offensive Rating", ylab="Defensive Rating")
cor(nba$f_tr, nba$n_rtg, use = "complete.obs")
barplot(nba$ts_percent)
library ( ggplot2 )
library ( tidyr )
boxplot(nba$ts_percent)
data_counts <- table(nba$playoffs, nba$team)
pie(data_counts)
#pie(nba$team)
boxplot(nba$e_fg_percent)
barplot(nba$e_fg_percent)
#pie(nba$e_fg_percent)
pie(data_counts)
barplot(data_counts)
#ggplot(data_counts)
data_counts_df <- as.data.frame(data_counts)
#ggplot(data = data_counts_df, aes(x = Var2, fill = Var1)) +
#geom_bar(position = "dodge") +
#labs(title = "",
#x = "",
#y = "") +
#theme_minimal()
ball <-table(nba$team, nba$tov_percent)
pie(ball)
#levels(factor(ball))
barplot(ball)
#pie(nba$playoffs)
barplot(nba$tov_percent)
#ggplot(nba$playoffs)
balll<- table(nba$playoffs, nba$team)
pie (balll, labels = nba$team,
main = " Diamonds by Cut ( Pie ) " ,
col = rainbow ( length ( nba$team ))
+theme_minimal())
library(dplyr)
top_5 <- head(balll, 5)
pie(top_5)
sorted_teamss <- sort(table(nba$team), decreasing = TRUE)
top_5 <- head(sorted_teamss, 5)
pie(top_5, main ="Teams that made the playoffs the most times" )
pct <- round (100 * top_5 / sum ( top_5 ) , 1)
lbls <- paste ( names ( top_5) , paste0 ( pct , " % " ))
pie(top_5, labels = lbls, main = "Teams that made the playoffs the most times
")
sd(nba$ts_percent)
sd(nba$f_tr)
sd(nba$e_fg_percent)
mean(nba$n_rtg)
net <-na.omit(nba$n_rtg)
mean(net)
plot(net)
median(net)
hist(net)
mode(nba$tov_percent)
median(nba$tov_percent)
hist(nba$tov_percent)
mean(nba$tov_percent)
median(nba$season)
mode(nba$team)
team <- table(nba$team)
pie(team)
basket <- read.csv("/Users/jeeya/Downloads/TeamSummaries_raw.csv")
mean(basket$age)
age <- na.omit(basket$age)
mean(age)
#hist(basket$w ~ basket$l)
lose <- na.omit(basket$l)
win <- na.omit(basket$w)
boxplot(win, lose, names = c("win","Lose"))
pw <- na.omit(basket$pw)
pl <- na.omit(basket$pl)
boxplot(pw, pl, names= c("p wins","p loses"))
hist(basket$mov, xlab = "margin of victory")
median(basket$sos)
sos <- na.omit(basket$sos)
median(sos)
mean(sos)
hist(sos)
sd(basket$srs)
srs <- na.omit(basket$srs)
sd(srs)
off <- na.omit(basket$o_rtg)
mean(off)
mean(basket$pace)
pace <-na.omit(basket$pace)
mean(pace)
median(pace)
orb <- na.omit(basket$orb_percent)
mode(orb)
mean(orb)
drb <- na.omit(basket$drb_percent)
boxplot(orb, drb, names = c("offensive rebound %","defensiverebound%"))
sd(basket$opp_e_fg_percent)
opp <- na.omit(basket$opp_e_fg_percent)
sd(opp)
hist(opp)
tov<- na.omit(basket$opp_tov_percent)
ft <- na.omit(basket$opp_ft_fga)
hist(tov)
median(tov)
hist(ft)
median(ft)
mean(basket$ft_fga)
fga <- na.omit(basket$ft_fga)
mean(fga)
hist(fga)
#medain(fga)
median(fga)
att <- na.omit(basket$attend_g)
hist(att, xlab= "game attandance")
median(att)
```

II. Regression
a. Multiple Linear Regression - Jolisia

Loading the Dataset
```{r}
library(tidyverse)
library(caret)

TeamData_raw <- read.csv("/Users/jeeya/Downloads/TeamSummaries_raw.csv")

TeamData_Clean <- TeamData_raw %>%
  filter(season >= 2021 & season <= 2026) %>%
  select(season, team, w, o_rtg, d_rtg, ts_percent) %>%
  drop_na()

head(TeamData_Clean)

str(TeamData_Clean)
summary(TeamData_Clean)
```

Correlation Matrix
```{r}
round(cor(TeamData_Clean[, c("w", "o_rtg", "d_rtg", "ts_percent")]), 4)
```

Multiple Linear Regression Plots
```{r}
model_mlr <- lm(w ~ o_rtg + d_rtg + ts_percent, data = TeamData_Clean)
summary(model_mlr)
plot(model_mlr$fitted.values,
     resid(model_mlr),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")

abline(h = 0, col = "red", lwd = 2)

hist(rstandard(model_mlr),
     col = "orange",
     main = "Histogram of Standardized Residuals",
     xlab = "Standardized Residuals")

qqnorm(rstandard(model_mlr))
qqline(rstandard(model_mlr), col = "red", lwd = 2)

res <- rstandard(model_mlr)
shapiro.test(res)
```

b. Polynomial Regression - Jolisia

Train-Test Split
```{r}
set.seed(123)

n <- nrow(TeamData_Clean)

train_idx <- sample(1:n, size = round(0.7 * n))

train <- TeamData_Clean[train_idx, ]
test <- TeamData_Clean[-train_idx, ]

model_Tr <- lm(w ~ o_rtg + d_rtg + ts_percent, data = train)

summary(model_Tr)
```

Train-Test Model Plot and Error Metrics
```{r}
plot(model_Tr$fitted.values,
     resid(model_Tr),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted - Training Model")

abline(h = 0, col = "blue", lwd = 2)

pred_test <- predict(model_Tr, newdata = test)

mse <- mean((test$w - pred_test)^2)
rmse <- sqrt(mse)
mae <- mean(abs(test$w - pred_test))

mse
rmse
mae
```

Polynomial Regression Plot
```{r}
x <- TeamData_Clean$o_rtg
y <- TeamData_Clean$w

fit_lin <- lm(w ~ o_rtg, data = TeamData_Clean)

fit_quad <- lm(w ~ poly(o_rtg, 2, raw = TRUE), data = TeamData_Clean)

fit_cub <- lm(w ~ poly(o_rtg, 3, raw = TRUE), data = TeamData_Clean)

summary(fit_lin)
summary(fit_quad)
summary(fit_cub)

xg <- seq(min(TeamData_Clean$o_rtg),
          max(TeamData_Clean$o_rtg),
          length.out = 200)

p_lin <- predict(fit_lin,
                 newdata = data.frame(o_rtg = xg))

p_quad <- predict(fit_quad,
                  newdata = data.frame(o_rtg = xg))

p_cub <- predict(fit_cub,
                 newdata = data.frame(o_rtg = xg))

plot(x, y,
     pch = 16,
     cex = 0.9,
     xlab = "Offensive Rating",
     ylab = "Wins",
     main = "Polynomial Regression: Wins vs Offensive Rating")

lines(xg, p_lin, lwd = 2, col = "blue")

lines(xg, p_quad, lwd = 2, lty = 2, col = "red")

lines(xg, p_cub, lwd = 2, lty = 3, col = "green4")

legend("topright",
       legend = c("Linear", "Quadratic", "Cubic"),
       lwd = 2,
       lty = c(1, 2, 3),
       col = c("blue", "red", "green4"),
       bty = "n")
```

AIC and BIC Comparison
```{r}
AIC(fit_lin, fit_quad, fit_cub)
BIC(fit_lin, fit_quad, fit_cub)
```

c. Principal Component Regression (PCR) - Ethne

Loading and Cleaning Data
```{r}
library(ISLR2)
library(ggplot2)
data <- read.csv("/Users/jeeya/Downloads/TeamSummaries_raw.csv")
teams_data <- na.omit(data[, c("mov","sos","pace","f_tr","x3p_ar","ts_percent","e_fg_percent","tov_percent","orb_percent","ft_fga","opp_e_fg_percent","opp_tov_percent","drb_percent","opp_ft_fga")])
```

PCR Split
```{r}
library(pls)
set.seed(123)
n <- nrow(teams_data)
train_idx <- sample(1:n, size = round(0.7*n))
train <- teams_data[train_idx, ]
test <- teams_data[-train_idx, ]
dim(train)
dim(test)
```

Fit PCR on Train Data
```{r}
pcr_train <- pcr(mov ~ .-mov, data = train, scale = TRUE, validation = "CV")
summary(pcr_train)
validationplot(pcr_train, val.type = "MSEP")
```

PCR on Test Data & RMSE
```{r}
pred_pcr <- predict(pcr_train, test, ncomp = 10)
RMSE <- function(y, yhat) sqrt(mean((y - yhat)^2))
test_rmse_pcr <- RMSE(test$mov, pred_pcr)
test_rmse_pcr
```

OLS Fit
```{r}
ols_fit <- lm(mov ~ .-mov, data = train)
summary(ols_fit)
```

d. Induction Tree - Ethne
```{r}
library(ISLR2)
library(rpart)
library(rpart.plot)
data <- read.csv("/Users/jeeya/Downloads/TeamSummaries_raw.csv")
data <- na.omit(data)
set.seed(123)
n <-nrow(data)
train_idx <- sample(1:n, 0.7 * n)
train <- data[train_idx, ]
test <- data[-train_idx, ]
tree_model <- rpart(w ~ pace + o_rtg + d_rtg + ts_percent + sos,
data = train, method = "anova",
control = rpart.control(cp = 0.01))
rpart.plot(tree_model, type = 4, extra = 1,
clip.right.labs = FALSE,
varlen = 0, faclen = 3)
printcp(tree_model)
plotcp(tree_model)
```

Fit Pruned Tree
```{r}
fit_pr <- prune(tree_model, cp = 0.017459)
rpart.plot(fit_pr, type = 4, extra = 1, fallen.leaves = TRUE,
digits = 3, varlen = 0, faclen = 3)
```

Predict and Compare
```{r}
pred <- predict(fit_pr, test)
head(data.frame(Actual = test$w, Predicted = pred))
rmse <- function(y, yhat) sqrt(mean((y-yhat)^2))
test_rmse_tree <- rmse(test$w, pred)
test_rmse_tree
```

III. Classification - Jeeya

Loading and Cleaning Data
``` {r}
library(tidyverse)
library(caret)
library(e1071)
library(rpart)
library(rpart.plot)

set.seed(123)
nba <- read.csv("/Users/jeeya/Downloads/TeamSummaries_raw.csv")

nba_clean <- nba %>%
  filter(lg == "NBA", season >= 2021, season <= 2026) %>%
  select(season, team, w, mov, srs, o_rtg, d_rtg, n_rtg, pace,
         ts_percent, e_fg_percent, tov_percent, orb_percent,
         drb_percent, opp_e_fg_percent, opp_tov_percent, ft_fga) %>%
  drop_na()

nba_clean$good_team <- ifelse(nba_clean$w > median(nba_clean$w, na.rm = TRUE), "Yes", "No")
nba_clean$good_team <- as.factor(nba_clean$good_team)

table(nba_clean$good_team)
```

Split Data into Training and Test Sets
```{r}
train_index <- createDataPartition(nba_clean$good_team, p = 0.70, list = FALSE)

train <- nba_clean[train_index, ]
test <- nba_clean[-train_index, ]

nrow(train)
nrow(test)
```

a. Logistic Regression
```{r}
model_log <- glm(good_team ~ mov + srs + o_rtg + d_rtg + n_rtg + pace +
                   ts_percent + e_fg_percent + tov_percent + orb_percent +
                   drb_percent + opp_e_fg_percent + opp_tov_percent + ft_fga,
                 data = train,
                 family = binomial)

summary(model_log)

prob_log <- predict(model_log, test, type = "response")

pred_log <- ifelse(prob_log > 0.5, "Yes", "No")
pred_log <- factor(pred_log, levels = levels(test$good_team))

conf_log <- table(Predicted = pred_log, Actual = test$good_team)
conf_log

accuracy_log <- mean(pred_log == test$good_team)
misclass_log <- 1 - accuracy_log

accuracy_log
misclass_log
```

b. Classification Tree
```{r}
model_tree <- rpart(good_team ~ mov + srs + o_rtg + d_rtg + n_rtg + pace +
                      ts_percent + e_fg_percent + tov_percent + orb_percent +
                      drb_percent + opp_e_fg_percent + opp_tov_percent + ft_fga,
                    data = train,
                    method = "class")

rpart.plot(model_tree)

pred_tree <- predict(model_tree, test, type = "class")

conf_tree <- table(Predicted = pred_tree, Actual = test$good_team)
conf_tree

accuracy_tree <- mean(pred_tree == test$good_team)
misclass_tree <- 1 - accuracy_tree

accuracy_tree
misclass_tree
```

c. Naive Bayes - Sophia
```{r}
model_nb <- naiveBayes(good_team ~ mov + srs + o_rtg + d_rtg + n_rtg + pace +
                         ts_percent + e_fg_percent + tov_percent + orb_percent +
                         drb_percent + opp_e_fg_percent + opp_tov_percent + ft_fga,
                       data = train)

model_nb

pred_nb <- predict(model_nb, test)

conf_nb <- table(Predicted = pred_nb, Actual = test$good_team)
conf_nb

accuracy_nb <- mean(pred_nb == test$good_team)
misclass_nb <- 1 - accuracy_nb

accuracy_nb
misclass_nb
```

d. Support Vector Machine (SVM) - Sophia
```{r}
model_svm <- svm(good_team ~ mov + srs + o_rtg + d_rtg + n_rtg + pace +
                   ts_percent + e_fg_percent + tov_percent + orb_percent +
                   drb_percent + opp_e_fg_percent + opp_tov_percent + ft_fga,
                 data = train,
                 kernel = "radial",
                 scale = TRUE)

model_svm

pred_svm <- predict(model_svm, test)

conf_svm <- table(Predicted = pred_svm, Actual = test$good_team)
conf_svm

accuracy_svm <- mean(pred_svm == test$good_team)
misclass_svm <- 1 - accuracy_svm

accuracy_svm
misclass_svm
```

Model Comparison
```{r}
comparison <- data.frame(
  Model = c("Logistic Regression", "Naive Bayes", "Classification Tree", "SVM"),
  Accuracy = c(accuracy_log, accuracy_nb, accuracy_tree, accuracy_svm),
  Misclassification_Rate = c(misclass_log, misclass_nb, misclass_tree, misclass_svm)
)

comparison
```

Visualizations
Bar Plot of Good vs Not Good Teams
```{r}
ggplot(nba_clean, aes(x = good_team, fill = good_team)) +
  geom_bar() +
  scale_fill_manual(values = c("Yes" = "orange", "No" = "black")) +
  labs(title = "Distribution of Good vs Not Good Teams",
       x = "Good Team",
       y = "Number of Teams") +
  theme_minimal()
```

Model Accuracy Comparison Plot
```{r}
ggplot(comparison, aes(x = Model, y = Accuracy, fill = Model)) +
  geom_col() +
  scale_fill_manual(values = c("steelblue", "orange", "purple", "darkred")) +
  labs(title = "Accuracy Comparison Across Classification Models",
       x = "Classification Model",
       y = "Accuracy") +
  ylim(0,1) +
  theme_minimal() +
  theme(legend.position = "none")
```

Wins by Good Team Classification
```{r}
ggplot(nba_clean, aes(x = good_team, y = w, fill = good_team)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Yes" = "orange", "No" = "black")) +
  labs(title = "Wins by Good Team Classification",
       x = "Good Team",
       y = "Wins") +
  theme_minimal()
```


IV. K Means Clustering - Ethne

Elbow Plot
```{r}
data <- read.csv("/Users/jeeya/Downloads/TeamSummaries_raw.csv")
data_omit <- na.omit(data)
tot.withinss<- vector(mode="numeric", length = 10)
set.seed(123)
for(i in 1:10){
teamsElbow <- kmeans(data_omit[, c("o_rtg", "d_rtg", "pace", "ts_percent",
"x3p_ar", "tov_percent", "orb_percent")], centers = i, nstart = 25)
tot.withinss[i] <- teamsElbow$tot.withinss}
plot(1:length(tot.withinss), tot.withinss,
type = "b", pch = 19,
main = "Elbow Plot",
ylab = "Total Within SS",
xlab = "Number of Centers (clusters)")
```

Cluster Plot
```{r}
library(cluster)
set.seed(123)
teamsCluster <- kmeans(x = data_omit[, c("o_rtg", "d_rtg","ts_percent",
"pace", "x3p_ar", "tov_percent", "orb_percent")],centers = 4, nstart = 25)
teamsCluster
clusplot(x=data_omit[, c("o_rtg", "d_rtg", "pace", "ts_percent", "x3p_ar",
"tov_percent", "orb_percent")],
teamsCluster$cluster,
main = "Cluster Plot",
color = TRUE,
shade = TRUE,
labels = 0,
lines = 0)
```