---
title: "HW2_ANLY3339"
author: "Jeeya Tanna"
date: "2026-03-02"
output: word_document
---
Step 1: Load the dataset
```{r}
library(ISLR2)
library(pls)
library(caret)

data(Hitters)
```
Step 2: Remove missing values and creat LogSalary
```{r}
Hitters2 <- na.omit(Hitters)
Hitters2$logSalary <- log(Hitters2$Salary)

nrow(Hitters2)
```
Number of observation after removing the missing values = 263

Step 3: Train-Test Split
```{r}
set.seed(123)

train_index <- createDataPartition(Hitters2$logSalary, p = 0.7, list = FALSE)

train <- Hitters2[train_index, ]
test  <- Hitters2[-train_index, ]

nrow(train)
nrow(test)
```
Training observations = 185
Test observations = 78

Step 4: Fit PCR on training set
```{r}
set.seed(123)

pcr_model <- pcr(logSalary ~ . - Salary,
                 data = train,
                 scale = TRUE,
                 validation = "CV")

summary(pcr_model)
validationplot(pcr_model, val.type = "MSEP")
```
Based on the CV plot and RMSEP table, I select c = 5 components because the prediction error reaches its minimum at that point.

In the summary(pcr_model) output:
X row = cumulative % variance explained in predictors.
logSalary row = cumulative % variance explained in the response.

With 1 component, X explains 38.56% of predictor variation and 51.80% of response variation.
With 3 components, X explains 71.01% of predictor variation and 53.20% of response variation.

Step 5: Evaluate PCR on Test set
```{r}
c <- 5
pcr_pred <- predict(pcr_model, test, ncomp = c)
pcr_rmse <- sqrt(mean((test$logSalary - pcr_pred)^2))
pcr_rmse
```

Step 6: Multiple Linear Regression
```{r}
mlr_model <- lm(logSalary ~ . - Salary, data = train)
mlr_pred <- predict(mlr_model, test)
mlr_rmse <- sqrt(mean((test$logSalary - mlr_pred)^2))
mlr_rmse
```

FINAL COMPARISON QUESTIONS

1. PCR Test RMSE = 0.7144598
MLR Test RMSE = 0.7126638
The model with lower RMSE, i.e. Multiple Linear Regression, performs better on the test set because it has a smaller prediction error.

2. Comparing R² on Training set
```{r}
# PCR R2
R2(predict(pcr_model, train, ncomp = c), train$logSalary)

# MLR R2
summary(mlr_model)$r.squared
```
On the training set, MLR has R² = 0.6164604 and PCR has R² = 0.5546967
MLR typically fits training data better because it has a higher R² value and it uses all predictors directly, whereas PCR compresses predictors into components.