---
title: "In-Class-Activity4"
author: "Jeeya Tanna"
date: "2026-03-04"
output: word_document
---

Step 1: Load dataset
```{r}
library(ISLR2)
library(splines)
data(Wage)
d <- na.omit(Wage[, c("wage", "age")])
```

Step 2: Split data into training and test sets
```{r}
n <- nrow(d)
train_idx <- sample(1:n, size = 0.7 * n)
train <- d[train_idx, ]
test  <- d[-train_idx, ]

cat("Train n =", nrow(train), "\n")
cat("Test n  =", nrow(test), "\n")

kn <- as.numeric(quantile(train$age, probs = c(0.25, 0.50, 0.75)))
cat("Knots:", kn, "\n")
```

Step 3: Fit quadratic polynomial regression to train set
```{r}
fit_poly <- lm(wage ~ poly(age, 2, raw = TRUE), data = train)
b <- coef(fit_poly)
cat(sprintf("\nQuadratic polynomial:\nwage_hat = %.4f + %.4f*age + %.4f*age^2\n", b[1], b[2], b[3]))
```

Step 4: Fit quadratic spline on train set
```{r}
fit_s2   <- lm(wage ~ bs(age, degree = 2, knots = kn), data = train)
b2 <- coef(fit_s2)
cat("\nQuadratic spline (B-spline basis):\n")
cat(sprintf("wage_hat = %.4f + %.4f*B1(age) + %.4f*B2(age) + %.4f*B3(age) + %.4f*B4(age) + %.4f*B5(age)\n",
            b2[1], b2[2], b2[3], b2[4], b2[5], b2[6]))
```

Step 5: Fit cubic spline on train set
```{r}
fit_s3   <- lm(wage ~ bs(age, degree = 3, knots = kn), data = train)
b3 <- coef(fit_s3)
cat("\nCubic spline (B-spline basis):\n")
cat(sprintf("wage_hat = %.4f + %.4f*B1(age) + %.4f*B2(age) + %.4f*B3(age) + %.4f*B4(age) + %.4f*B5(age) + %.4f*B6(age)\n",
            b3[1], b3[2], b3[3], b3[4], b3[5], b3[6], b3[7]))
```

Step 6: Fit simple linear regression model on train set
```{r}
fit_lin  <- lm(wage ~ age, data = train)
bl <- coef(fit_lin)
cat(sprintf("\nSimple linear:\nwage_hat = %.4f + %.4f*age\n", bl[1], bl[2]))
```

Summary tables:
```{r}
model_stats <- function(m) {
  s <- summary(m)
  fp <- pf(s$fstatistic[1], s$fstatistic[2], s$fstatistic[3], lower.tail = FALSE)
  c(R2 = s$r.squared, Adj_R2 = s$adj.r.squared, RSE = s$sigma, F_pvalue = fp, BIC = BIC(m))
}

cat("\n--- Coefficient significance (use p-values) ---\n")
print(summary(fit_poly)$coefficients)
print(summary(fit_s2)$coefficients)
print(summary(fit_s3)$coefficients)
print(summary(fit_lin)$coefficients)

cat("\n--- Model stats on TRAIN ---\n")
stats <- rbind(
  Polynomial = model_stats(fit_poly),
  QuadSpline = model_stats(fit_s2),
  CubicSpline = model_stats(fit_s3),
  Linear = model_stats(fit_lin)
)
print(stats)
```

Step 7: Evaluate models on train set with BIC
```{r}
cat("\nBIC ranking (train):\n")
print(sort(c(
  Polynomial = BIC(fit_poly),
  QuadSpline = BIC(fit_s2),
  CubicSpline = BIC(fit_s3),
  Linear = BIC(fit_lin)
)))
```

Step 8: Evaluate models on test set
```{r}
rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))

pred_poly <- predict(fit_poly, newdata = test)
pred_s2   <- predict(fit_s2,   newdata = test)
pred_s3   <- predict(fit_s3,   newdata = test)
pred_lin  <- predict(fit_lin,  newdata = test)

rmse_vals <- c(
  Polynomial = rmse(test$wage, pred_poly),
  QuadSpline = rmse(test$wage, pred_s2),
  CubicSpline = rmse(test$wage, pred_s3),
  Linear = rmse(test$wage, pred_lin)
)
cat("\nRMSE ranking (test):\n")
print(sort(rmse_vals))
```

Step 9: Visualize fits on training data
```{r}
age_seq <- seq(min(train$age), max(train$age), length.out = 300)

plot(train$age, train$wage, pch = 19, col = "grey70",
     xlab = "Age", ylab = "Wage", main = "Wage ~ Age (Train)")
lines(age_seq, predict(fit_lin,  data.frame(age = age_seq)), col = "black", lwd = 2, lty = 2)
lines(age_seq, predict(fit_poly, data.frame(age = age_seq)), col = "red", lwd = 2)
lines(age_seq, predict(fit_s2,   data.frame(age = age_seq)), col = "blue", lwd = 2)
lines(age_seq, predict(fit_s3,   data.frame(age = age_seq)), col = "orange", lwd = 2)
legend("topright",
       legend = c("Linear", "Polynomial (deg 2)", "Quadratic spline", "Cubic spline"),
       col = c("black", "red", "blue", "orange"),
       lty = c(2, 1, 1, 1), lwd = 2, bty = "n")
```
